---
title: "Discussion 1 Introduction & Summarizing Data"
author: "Stats 67 Teaching Team"
format: html
---

# Load Packages

```{r}
library(datasets)
library(tidyverse)
```


# About R

Basic functions and language

```{r}
print('hello woRld')
birth_years <- c(1996, 2001, 1999)
print(birth_years)
```

# Dataframe in R

```{r}
data.frame(person = c('Jack', 'Allen', 'Lisa'), birth_years = birth_years)

?data.frame
```

You could also check the use of a function or a package by searching in the help panel.

# Tidyverse

After function names do not leave any spaces.

Before and after operators (e.g. <-, =) leave spaces.

Put a space after a comma, not before.

Object names are all lower case, with words separated by an underscore.


# Markdown and Quarto

word

_word_

__word__

~~word~~

If you need R variables in text, you can do this: the birth_years are:
`r birth_years`

# Load mtcars Dataset

```{r}
mtcars <- mtcars %>% 
  mutate(am = as.factor(am))
```

# Dataset

## Description given by the author

```{r}
?mtcars
```

# First Look

Take a glimpse at the data.

1.  How many rows and columns are there in the "mtcars" dataset?
2.  Any categorical variables?
3.  Any numeric variables?

```{r}
glimpse(mtcars)
```

# First and last few lines

```{r}
head(mtcars)
```

```{r}
tail(mtcars)
```


# Summarizing Categorical Data

## Count by Transmission Types (am)

How many samples are there in each transmission type?

```{r}
count(mtcars, am)
```

## Bar plots

### Bar plot with counts

```{r}
ggplot(mtcars) +
  aes(x = am) +
  geom_bar()
```

# Summarizing Numeric Data

## A few basic statistics

### Mean

What is the meaning of "Mean"?

```{r}
summarize(mtcars, mean(mpg))
```

### Median

What is the meaning of "Median"?

```{r}
summarize(mtcars, median(mpg))
```

### Quantiles

The 3rd Quantile of mpg:

```{r}
summarize(mtcars, quantile(mpg, 0.75))
```

What about the 1st Quantile? What is the meaning of "Quantile"?

### Standard Deviation and Variance

What are the meanings of "Standard Deviation" and "Variance"?

```{r}
summarize(mtcars, sd(mpg))
```

```{r}
summarize(mtcars, var(mpg))
```

## Histogram

```{r}
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(bins = 8)
```

## Boxplot of Miles per Gallon (mpg)

On the boxplot, where is the median? Q1 and Q3? Interquartile Range (IQR)? Whiskers?

```{r}
ggplot(mtcars,
       aes(y = mpg)) + 
  scale_x_discrete( ) +
  geom_boxplot()
```

# Summarizing More Than One Variable

## Boxplot of Miles per Gallon (mpg) by Transmission Types (am)

How many variables are represented in this plot?

Are they numeric or categorical?

```{r}
ggplot(mtcars, aes(x = am, y = mpg)) +
  geom_boxplot()
```

## Scatterplot of Miles per Gallon (mpg) and Horse Power (hp)

How many variables are represented in this plot?

Are they numeric or categorical?

```{r}
ggplot(mtcars, aes(x = mpg,
                 y = hp)) +
  geom_point()
```






