```{r}
#| message: false
#| warning: false
library(tidyverse)
```

# Discrete Random Variable

Let $X$ be the number you get when rolling an unfair die.

| $x$      | 1   | 2   | 3    | 4   | 5   | 6    |
|:-------|:----|:----|:-----|:----|:----|:-----|
| $P(X=x)$ | 0.2 | 0.1 | 0.15 | 0.1 | 0.3 | 0.15 |

## Probability Mass Function (pmf)

$P(X=1)=f(1)=0.2$

$P(X=2)=f(2)=0.1$

$P(X=3)=f(3)=0.15$

$P(X=4)=f(4)=0.1$

$P(X=5)=f(5)=0.3$

$P(X=6)=f(6)=0.15$

```{r, echo=F}
x <- 1:6
y <- c(0.2, 0.1, 0.15, 0.1, 0.3, 0.15)

data <- data.frame(x = x, y = y)

data %>% 
  ggplot() +
  aes(x = x, y = y) +
  geom_segment(data = data, 
               aes(x = x, y = 0, 
                   xend = x, yend = y)) +
  labs(x = expression(paste("x")),
       y = expression(paste("f(x)"))) +
  theme_bw() +
  theme(text = element_text(size = 20)) + 
  scale_x_continuous(breaks = pretty(data$x, n = 6))
```

## Cumulative Distribution Function (cdf)

| $x$      | 1   | 2   | 3    | 4   | 5   | 6    |
|:-------|:----|:----|:-----|:----|:----|:-----|
| $P(X\leq x)$ | 0.2 | 0.3 | ? | ? | ? | 1 |

$P(X\leq1)=F(1)=0.2$

$P(X\leq2)=F(2)=0.3$

...

$P(X\leq6)=F(6)=1$

_Exercise 1: Fill in the cdf table using the pmf table._

_Exercise 2: If we were only given the completed cdf table, are we able to know the pmf of the random variable?_

_Exercise 3: What are $F(0), F(2.5), F(10)$?_

```{r, echo=F}
x <- 1:6
y <- cumsum(c(0.2, 0.1, 0.15, 0.1, 0.3, 0.15))
xend <- c(2, 3, 4, 5, 6, 6)

data <- data.frame(x = x,
                   y = y,
                   xend = xend)

ggplot(data = data,
       aes(x = x, y = y, xend = xend, yend = y)) +
  geom_point() +
  geom_segment() +
  scale_x_continuous(breaks=1:6,
                     limits = c(1, 6)) +
  scale_y_continuous(limits = c(-.1, 1.1)) +
  labs(x = "x",
      y = expression(paste("F(x)")))+
  theme_bw() +
  theme(text = element_text(size = 20)) 
```

## Expectation

| $x$    | 1   | 2   | 3    | 4   | 5   | 6    |
|:-----|:----|:----|:-----|:----|:----|:-----|
| $f(x)$ | 0.2 | 0.1 | 0.15 | 0.1 | 0.3 | 0.15 |

$E[X]=\sum_{x\in S}xf(x)=1\times0.2  +2\times 0.1 +3\times  0.15 +4\times 0.1 +5\times  0.3 +6\times  0.15=3.65$

```{r}
x <- 1:6
f_x <- c(0.2, 0.1, 0.15, 0.1, 0.3, 0.15)
mu <- sum(x * f_x)
mu
```

## Variance

$Var(X)=E[(X-E[X])^2]=E[X^2]-(E[X])^2$, where

$E[X^2]=\sum_{x\in S}x^2f(x)=1^2\times0.2  +2^2\times 0.1 +3^2\times  0.15 +4^2\times 0.1 +5^2\times  0.3 +6^2\times  0.15=16.45$

and

$(E[X])^2=3.65^2=13.3225$

```{r}
mu_2 <- sum(x^2 * f_x)
mu_2 - mu^2
```

## Linear Combination of Random Variables

$E(aX+bY)=a\cdot E(X)+b\cdot E(Y)$

$Var(aX+bY)=a^2\cdot Var(X)+b^2\cdot Var(Y)$ if $X$ and $Y$ are independent.


# Discrete Distributions

## Bernoulli Distribution

The Bernoulli distribution represents the outcome (success/failure) of an event with success probability of $\pi$.

$X\sim Bernoulli(\pi)$

| x    | 0       | 1     |
|:-----|:--------|:------|
| f(x) | $1-\pi$ | $\pi$ |

pmf: $f(x)=\pi^x(1-\pi)^{1-x},\ x\in \{0,1\}$


$\mu=E(X)=\pi$

$\sigma^2=Var(X)=\pi(1-\pi)$

_Example: From a deck of 52 cards, randomly pick a card, let the event be that the suit of the card picked is heart._

## Geometric Distribution

The Geometric distribution represents the number of failures before the first success when running independent trials with success probability $p$.

$X\sim Geometric(p)$

pmf: $f(x)=(1-p)^xp\ ,x=0,1,2,...$

$\mu=E(X)=\frac{1-\pi}{\pi}$

$\sigma^2=Var(X)=\frac{1-\pi}{\pi^2}$

```{r, echo=F, warning=F}
# Parameters
p_values <- c(0.1, 0.3, 0.5)  # Probability values
x <- 0:10  # Range of x values (number of failures before the first success)

# Create a data frame for ggplot
data <- data.frame(
  x = rep(x, times = length(p_values)),
  p = unlist(lapply(p_values, function(p) dgeom(x, prob = p))),
  prob = rep(p_values, each = length(x))
)

# Plot using ggplot2
ggplot(data, aes(x = x, y = p, color = factor(prob), group = prob)) +
  geom_line(size=1) +             # Add lines
  geom_point(size=2) +            # Add points
  scale_color_manual(
    values = c("red", "blue", "green"),
    name = "p"
  ) + 
  scale_x_continuous(breaks=0:10) +
  labs(
    title = "PMF of Geometric Distribution",
    x = "Number of Failures (x)",
    y = "Probability"
  ) +
  theme_minimal(base_size = 16)  # Increase overall font size
```

_Exercise: From a deck of 52 cards, randomly pick a card, if the suit of the card picked is not heart, then replace the card into the deck, shuffle the deck and pick a new card. Repeat the procedure until the first heart is picked. What is the probability that the number of trials before the last one equals to 3? What is the probability that the number of trials before the last one is smaller or equal to 3?_

```{r, include=F}
# f(3)
dgeom(x = 3, prob = 0.25)

# Calculate F(3) by f(0) + f(1) + f(2) + f(3)
dgeom(x = 0, prob = 0.25) + dgeom(x = 1, prob = 0.25)  + dgeom(x = 2, prob = 0.25) + dgeom(x=3, prob=0.25)
sum(dgeom(x=0:3, prob=0.25))

# Directly calculate F(3)
pgeom(q = 3, prob = 0.25) 
```



### Exercises from [the book](https://www.openintro.org/go/?id=os4_for_screen_readers&referrer=/book/os/index.php)

4.14 on page 166

## Binomial Distribution

The Binomial distribution represents the number of successes out of $n$ independent trials with success probability $\pi$.

$X\sim Binomial(n,\pi)$

pmf: $f(x)={n \choose x}\pi^x(1-\pi)^{n-x},\ x\in \{0,1,2,...,n\}$

$E(X)=n\pi$

$Var(X)=n\pi(1-\pi)$

```{r, echo=F, warning=F}
# Binomial, fixed n=10
n <- 10  # Number of trials
p_values <- c(0.2, 0.5, 0.8)  # Probability of success
x <- 0:n  # Range of x values (number of successes)

# Create a data frame for ggplot
data <- data.frame(
  x = rep(x, times = length(p_values)),
  p = unlist(lapply(p_values, function(p) dbinom(x, size = n, prob = p))),
  prob = rep(p_values, each = length(x))
)

# Plot using ggplot2
ggplot(data, aes(x = x, y = p, color = factor(prob), group = prob)) +
  geom_line(size = 1) +             # Add lines
  geom_point(size = 2) +            # Add points
  scale_color_manual(
    values = c("red", "blue", "green"),
    name = "p"
  ) + 
  scale_x_continuous(breaks = 0:n) +
  labs(
    title = "PMF of Binomial Distribution, fixed n=10",
    x = "Number of Successes (x)",
    y = "Probability"
  ) +
  theme_minimal(base_size = 16)     # Increase overall font size
```

```{r, warning=F, echo=F}
p <- 0.5  # Fixed probability of success
n_values <- c(10, 20, 40)  # Different values for number of trials
x_values <- lapply(n_values, function(n) 0:n)  # Range of x values for each n

# Create a data frame for ggplot
data <- data.frame(
  x = unlist(x_values),
  p = unlist(lapply(seq_along(n_values), function(i) dbinom(x_values[[i]], size = n_values[i], prob = p))),
  n = factor(rep(n_values, times = sapply(x_values, length)))  # Add n as a factor
)

# Plot using ggplot2
ggplot(data, aes(x = x, y = p, color = n, group = n)) +
  geom_line(size = 1) +             # Add lines
  geom_point(size = 2) +            # Add points
  scale_color_manual(
    values = c("red", "blue", "green"),
    name = "n"
  ) + 
  labs(
    title = "PMF of Binomial Distribution, fixed p=0.5",
    x = "Number of Successes (x)",
    y = "Probability"
  ) +
  theme_minimal(base_size = 16)     # Increase overall font size
```

_Exercise: From a deck of 52 cards, randomly pick a card and record the suit, then replace the card into the deck, shuffle the deck, pick a new card and record the suit. Repeat the procedure 10 times. What is the probability that 8 of the 10 suits recorded are hearts? What is the probability that at least 8 of the 10 suits recorded are hearts?_

```{r, include=F}
# f(8)
dbinom(x = 8, size = 10, prob = 0.25)

# f(8) + f(9) + f(10)
dbinom(x = 8, size = 10, prob = 0.25) + dbinom(x = 9, size = 10, prob = 0.25) + dbinom(x = 10, size = 10, prob = 0.25)
sum(dbinom(x = 8:10, size = 10, prob = 0.25))

# 1 - F(7)
1 - pbinom(q = 7, size = 10, prob = 0.25)
```

### Exercises from [the book](https://www.openintro.org/go/?id=os4_for_screen_readers&referrer=/book/os/index.php)

4.22 on page 176

## Poisson Distribution

The Poisson distribution can represent the number of events occurring within a fixed period of time.

$X\sim Poisson(\lambda)$

pmf: $f(x)=\frac{\lambda^x}{x!}e^{-\lambda},\ x=0,1,2,...$

$E(X)=\lambda$

$Var(X)=\lambda$

```{r, echo=F, warning=F}
# Parameters for Poisson distribution
lambda_values <- c(1, 2, 5, 10)  # Lambda values
x <- 0:20  # Range of x values

# Create a data frame for ggplot
data <- data.frame(
  x = rep(x, times = length(lambda_values)),
  p = unlist(lapply(lambda_values, function(lambda) dpois(x, lambda = lambda))),
  lambda = rep(lambda_values, each = length(x))
)

# Plot using ggplot2
ggplot(data, aes(x = x, y = p, color = factor(lambda), group = lambda)) +
  geom_line(size = 1) +             # Add lines
  geom_point(size = 2) +            # Add points
  scale_color_manual(
    values = c("red", "blue", "green", "purple"),
    name = expression(lambda)
  ) + 
  scale_x_continuous(breaks = 0:20) +
  labs(
    title = "PMF of Poisson Distribution",
    x = "Number of Events (x)",
    y = "Probability"
  ) +
  theme_minimal(base_size = 16)     # Increase overall font size
```

_Exercise: It is estimated that one bus arrives at the stop every 6 minuts. What is the probability that 10 buses will arrive at the stop during the next hour? What is the probability that the number of buses that arrive next hour is between 4 and 15 (both included)?_
```{r, include=F}
lambda = 60 / 6

# f(10)
dpois(x = 10, lambda = lambda)

# f(4) + f(5) + ... + f(15)
sum(dpois(x = 4:15, lambda = lambda))

# F(15) - F(3)
ppois(q = 15, lambda = lambda) - ppois(q = 3, lambda = lambda)
```



### Exercises from [the book](https://www.openintro.org/go/?id=os4_for_screen_readers&referrer=/book/os/index.php)

4.32 on page 185
