---
title: "7-28"
format: html
---

## 95% Confidence Interval

```{r}
xbar1 <- 16.12
xbar2 <- 19.85
point_estimate <- xbar1-xbar2
point_estimate
```

```{r}
s1 <- 3.58
s2 <- 4.51
n1 <- 26
n2 <- 26
```


```{r}
critical_value <- qt(0.975, df = 25) 
critical_value
```

```{r}
se <- sqrt((s1^2)/n1 + (s2^2)/n2)
se
```

```{r}
point_estimate - critical_value*se
point_estimate + critical_value*se
```
Looking at this interval with only negative values, I am confident that the average mpg for automatic cars is lower than manual cars.

## Hypothesis Testing

$$H_0:\mu_1 - \mu_2 = 0$$
$$H_A:\mu_1 - \mu_2 \neq 0$$

```{r}
(point_estimate - 0) / se
```

```{r}
pt((point_estimate - 0) / se, df = 25) *2 
```
The p-value of `r pt((point_estimate - 0) / se, df = 25) *2` is less than 0.05 we reject the null hypothesis and conclude that there is a significant difference between average mpg of automatic cars and manual cars. Considering that automatic cars have a lower mpg we can conclude also that automatic cars have significantly lower mpg than manual cars.
