```{r, echo=F, message=F}
library(tidyverse)
```


# Writing and Testing Hypotheses

## Research Question

A research question is a clear, focused, concise, complex, and arguable question around which you center your research. For example:

> "Does a new teaching mehthod improve student test scores compared to traditional methods?"

## Hypotheses

From the research question, we can derive two hypotheses:

- Null Hypothesis ($H_0$): The new teaching method has no effect on student test scores.
- Alternative Hypothesis ($H_1$ or $H_a$): The new teaching method improves student test scores.

## Hypothesis Testing Procedure

The procedure to test the above hypothesis could involve the following steps:

1. **Set hypotheses**: Set null and alternative hypotheses.
2. **Collection of data**: Collect test scores before and after implementing new teaching methods.
3. **Statistical test**: Use an appropriate statistical test (like a t-test) to compare means.
4. **Decision**: Reject $H_0$ if there is evidence to reject the null hypothesis, else fail to reject $H_0$.

# P-value
## How do we know when to reject or fail to reject the null hypothesis?

In order to determine whether we will reject or fail to reject our null hypothesis, we can look at our p-value which helps us determine the significance of our results. So, if

$p \leq \alpha$, we reject the null hypothesis.
$p > \alpha$, we fail to reject the null hypothesis.

$\alpha$ represents our significance level (threshold for deciding whether to reject the null hypothesis).


# Central Limit Theorem

The Central Limit Theorem (CLT) states that, given a sufficiently large sample size, the sampling distribution of the mean will be normally distributed regardless of the original distribution of the data.

Important things to note:

1. Samples must be independent of each other

2. Identically Distributed: Samples should be drawn from the same population distribution

## Example

Here, we will demonstrate the CLT by simulating the roll of a die, and showing that the means of the rolls tend to a normal distribution.

```{r clt-example, echo=TRUE}
set.seed(123)
dice_rolls <- replicate(10000, mean(sample(1:6, size = 30, replace = TRUE)))
hist(dice_rolls, breaks = 30, main="Demonstration of CLT with Dice Rolls", xlab="Mean of 30 dice rolls")
```

# Confidence Interval for a Single Proportion

A confidence interval for a single proportion estimates the proportion of a population that has a particular attribute.

## Example

Suppose we want to estimate the proportion of people who approve of a new policy. In a random sample of 100 people, 60 approved the policy.

```{r ci-single-proportion, echo=TRUE}
# Calculate the confidence interval
n <- 100  # Sample size
p <- 0.6  # Sample proportion

# Confidence interval
library(binom)
binom.confint(n * p, n, conf.level = 0.95, methods="wilson")
```

# Hypothesis Testing for a Single Proportion

To test a hypothesis about a single proportion, we use a significance test such as the z-test for proportions.

## Example

Testing if the true proportion of voters who support a candidate is different from 50% based on a sample of 100 voters where 58 voters expressed their support.

```{r}
# Hypothesis test variables from the previous chunk
pi_0 <- 0.5  # Hypothesized proportion
p_hat <- 0.58  # Observed proportion
n <- 100  # Sample size
SE <- sqrt(p_hat * (1 - p_hat) / n)

# 95% Confidence Interval for the observed proportion
alpha <- 0.05
z_alpha <- qnorm(1 - alpha / 2)
CI_lower <- p_hat - z_alpha * SE
CI_upper <- p_hat + z_alpha * SE

# Normal Distribution based on observed proportion and SE
x <- seq(0.3, 0.7, length = 1000)
y <- dnorm(x, mean = p_hat, sd = SE)

# PDF
plot(x, y, type = "l", lwd = 2, col = "blue", main = "Distribution of the Observed Proportion", xlab = "Proportion", ylab = "Density")
abline(v = pi_0, col = "red", lwd = 2, lty = 2)

# Confidence Interval
abline(v = CI_lower, col = "green", lwd = 2, lty = 3)
abline(v = CI_upper, col = "green", lwd = 2, lty = 3)

# Adding legends
legend("topleft", legend=c("PDF", "Observed Proportion", "95% CI Lower", "95% CI Upper"), 
       col=c("blue", "red", "green", "green"), lty = c(1, 2, 3, 3), cex=0.8)


```

